---
title: "CO2 emissions. An analysis using Dynamic Factor Models"
description: |
  In this post we will use DFM to investigate  
  the CO2 emissions produced over the last 100 years by 15 countries of different geographical and
  economical areas. To estimate the models we will use the R package MARSS.
categories:
  - Econometrics
  - Time series
  - R
  - Environment
preview: Preview.jpg
author:
  - name: Alessandro Ghiretti 
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
code_folding: true
bibliography: references.bib
nocite: | 
  @durbin2012time, @harvey1989time, @dlmharvey, @MARSS
---

```{r setup, include=FALSE}
```
# Introduction
Carbon dioxide emissions are the primary driver of global climate change and
it is widely recognized that to avoid the worst impacts of climate change, 
the world needs to urgently reduce emissions.
Prior to the Industrial Revolution, emissions were very low and growth in emissions 
was still relatively slow until the mid-20th century. For example, in 1950 the world emitted 
just over 5 billion tonnes of (CO2), about the same as the US, or half of China’s 
annual emissions today. By 1990 this had quadrupled to 22 billion tonnes. 
Emissions have continued to grow rapidly and we now emit over 36 billion tonnes each year.
Emissions growth has slowed over the last few years, but they have yet to reach their peak.

The purpose of this post is to use dynamic factor models and try to identify 
hidden common trends between the three main geographical areas of EU, North America 
and Asia that can be used to describe the evolution of the CO2 emissions.
We will consider CO2 emissions from 15 of the main Global Economies.To fit our 
model we will use the R package MARSS


# Analysis 
## Data Description
The data that we consider available at the following link:

https://raw.githubusercontent.com/owid/co2-data/master/owid-co2-data.csv

The data contains the CO2 emissions, the GPD and other relevant variables for all the
world countries starting from 1715 to 1018.
In our analysis we will consider data starting from 1900 to 2018 for the following
countries: Italy, United Kingdom, France, Germany, Spain, United States, Canada,
Japan, India, China, Argentina, Brazil, Peru, New Zealand and Australia
 


```{r, echo = FALSE, messages = FALSE, warnings = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(MARSS)
library(formattable)
library(patchwork)
library(plotly)
library(reshape)
library(RColorBrewer)
```

## Exploratory Data Analysis

We start the analysis by loading the data into R. In my case the data in in the
project folder

```{r, echo = TRUE}
# Read data
co2 <- read.csv("owid-co2-data.csv")
```

We proceed by filtering out the countries all the variables that are not of interest
to us. In addition, as it might be useful to compare the results at continent
level we add the variable zone which contains the continent of all the countries 
retained in the analysis.

```{r, echo = TRUE}
# Filter countries to include in the analysis
countries <- c("Italy","United Kingdom","France","Germany","Spain","United States", 
               "Canada","Japan", "India", "China",  "Argentina",
               "Brazil", "Peru", "New Zealand", "Australia")

# Define geographical zones 
zone <- c(rep("Europe",5), rep("North America", 2), rep("Asia",2),
          rep("South America",3), rep("Oceania",2))

# Add geographical zones to the data
co2 <-co2 %>% 
  mutate(zone = ifelse(country %in% 
    c("Italy","United Kingdom","France","Germany","Spain"),
    "Europe",ifelse( country %in% c ("United States","Canada"), "North America",
    ifelse(country %in% c("Japan", "India", "China"), "Asia",
    ifelse(country %in% c("Argentina","Brazil", "Peru"),"South America", "Oceania")))))

# Filter the data 
dat <- co2 %>% 
  # filter the countries and the year
  filter(country %in% countries &
         year >= 1910) %>% 
  # select columns
  select(iso_code, country, year, zone, co2)
```

Now we are ready to analyze our data. Lets start by  
looking at the temporal structure of the CO2 emissions for the different countries.

```{r, echo = TRUE, code_folding=TRUE}
nb.cols <- length(countries)
mycolor <- mycolors <- colorRampPalette(brewer.pal(9, "YlGnBu"))(nb.cols)
# Co2 Emissions Plot
# Plot co2 emissions
p1 <- dat %>% ggplot(aes(x = year, y = co2, color = country)) +
  geom_line() +
  scale_color_manual(values = mycolor) +
  ggtitle("CO2 emissions time series")+
  ylab("CO2 emissions") +
  theme_minimal()

ggplotly(p1)  
```

As we immediately see there are big differences in the CO2 emissions. 
As expected USA is the country with the highest CO2 emissions for the entire period,
followed by China, whose emissions show an impressive positive trend starting from 1975.
Germany, Japan and India represent the other four countries with the highest CO2
emissions over the period considered. Since there are great differences between the series,
in order to reduce the discrepancies it might be useful to take a look at the $log$ transformed
series.


```{r}
p1 <- dat %>% ggplot(aes(x = year, y = log(co2), color = country)) +
  geom_line() +
  scale_color_manual(values = mycolor) +
  ggtitle("log(CO2) time series")+
  ylab("CO2 emissions") +
  theme_minimal()

ggplotly(p1)  
```

Now that we can look at the whole picture in more detail we immediately note that
over the period considered the CO2 emissions have increased steadily for all of the countries. 
Moreover, Peru and New Zealand emerge clearly as the countries with the lowest emissions.
Moreover, an interesting pattern is observed around 1945 where many countries 
exhibit severe downward spikes.


```{r, echo = TRUE, code_folding=TRUE}
 p1 <- dat %>% ggplot(aes(x = zone, y = co2, color = country)) +
  geom_boxplot() +
   scale_color_manual(values = mycolor) +
  ggtitle("CO2 concentration boxplot")+
  ylab("CO2 concentration") +
  theme_minimal()

 p1

```

When we consider the aggregated data over the entire period the differences in
the CO2 emission between countries are even more evident.
Oceania and South America are the zones with the lowest and less variable
registered emissions. In Europe the situation is more heterogeneous, where despite
emissions appear on average pretty low there is more variability between the countries.
Overall, with the  exception of USA and China, we observe
that the variability and the average value of the CO2 emissions are similar
between the countries that belong to the same geographical zone.

```{r, echo = TRUE, code_folding = TRUE}
 p1 <- dat %>% 
  group_by(year,zone) %>% summarise(co2 = sum(co2)) %>% 
  ggplot() +
  geom_line(aes(x = year, y = co2, color = zone),
            size = 1.2, alpha = 0.6) +
   scale_color_brewer(palette = "Set2") +
  ggtitle("CO2 total concentration by area")+
  ylab("CO2 concentration") +
  theme_minimal()

ggplotly(p1)
```


To conclude our exploratory analysis we compute and examine the correlation of the 
CO2 emissions between the different countries. 
It appears that in general there is high positive correlation between the 
CO2 levels observed in the different countries and apparently this correlation
is not strictly related to the geographical area.

```{r}
m <- dat %>% 
  select(country,year,co2) %>% 
  pivot_wider( names_from  = country, values_from = co2) %>% 
  select(-c("year")) %>% 
  cor() %>% melt()
  
ggplot(m, aes(x = X1, y = X2, fill = value), alpha = 0.4) +
  geom_tile() +
  scale_fill_gradient(low = "pink", high = "royalblue")+
  xlab("") + ylab("") +
  theme_minimal()+  theme( panel.background = element_blank(),
                          axis.text.x = element_text(size = 7, angle = 90))
```




## Model specification and estimation
Having explored our data we proceed to specify the model.
In this case we will use a dynamic factor model to identify common trends
between the CO2 emissions of the different countries.
Compared to VAR,  in dynamic factor models require a lower number of parameters
need to be estimated, and therefore they result particularly useful when
then number os series considered is high.
The genera idea behind DFM is that the behavior of $k$ time series can be described
by a set of $m$ common states where in general $m$ is much smaller than $k$.
After the $m$ states are obtained they can be used for different purposes, for 
example to identify common structures in the series, such as trends, cyclical
or seasonal patterns or as proxy variables that resumes global structures of the
underlying series. Following Harvey (referenze)
and MARSS we cosnider the following Dynamic Factor model (DFM).


$$
\begin{aligned}

\begin{pmatrix}
y_{1} \\
y_{2} \\
\vdots \\
y_{m+k}
\end{pmatrix}_{t}

&= 

\begin{pmatrix}
b_{11} & 0 & \cdots & 0 \\
b_{21} & b_{22} & \cdots & 0 \\
\vdots& \vdots & \ddots & \vdots \\
b_{k1} & b_{k2} & \cdots & b_{km}
\end{pmatrix}

\begin{pmatrix}
x_{1} \\
x_{2} \\
\vdots \\
x_{m}
\end{pmatrix}_{t}

+

\begin{pmatrix}
w_{1} \\
w_{2} \\
\vdots \\
w_{m}
\end{pmatrix}_{t} \\






\begin{pmatrix}
x_{1} \\
x_{2} \\
\vdots \\
x_{m}
\end{pmatrix}_{t}

&= 

\begin{pmatrix}
1 & 0 & \cdots & 0 \\
0 & 1 & \cdots & 0 \\
\vdots& \vdots & \ddots & \vdots \\
0 & 0 & \cdots & 1
\end{pmatrix}

\begin{pmatrix}
x_{1} \\
x_{2} \\
\vdots \\
x_{m}
\end{pmatrix}_{t-1}

+

\begin{pmatrix}
v_{1} \\
v_{2} \\
\vdots \\
v_{m+k}
\end{pmatrix}_{t}
\end{aligned}
$$

According to model (ref) the $k$ CO2 emission time series are related to $m$ common hidden trends, 
each one described by a random walk and possibly correlated to the others.
The relevance that each state has on the observed series is expressed by the corresponding
factor loading $b_{jq}$.
According to the explanatory analysis we fit a model with four states and an unrestricted
covariance matrix. As it is generally suggested when fitting this kind of models the observations
are standardized. In the following  we reshape our data in a $N \times T$ matrix and
proceed to fit the model

```{r, echo = TRUE}
# row: country column: year
y <- dat %>% 
  select(-c("zone")) %>% 
  pivot_wider(names_from = year, values_from = co2) %>% 
  arrange(iso_code)

# convert the data in matrix
y = as.matrix(y[,3:ncol(y)])

# set parameters for optimizer
cntl.list <- list(minit = 200, maxit = 5000, allow.degen = FALSE)

#  Define values for grid search
# define covariance structures for the states
R.structure <- "unconstrained"
m <- 4

model.list <- list(m = m, R = R.structure)
# fit model to standardized data
mod.me <- MARSS(y, model = model.list, z.score = TRUE, form = "dfa",
                control = cntl.list, silent = TRUE)
```

### Estimated trends
Once we have estimated our model we can extract and plot the states of the system.
Since there exists many possible solutions to model (ref), we apply the varimax
rotation of the factors and of the trends. (referenza)

```{r, echo = TRUE}
# perform rotation of the states before plotting
# get the inverse of the rotation matrix
Z.est <- coef(mod.me, type = "matrix")$Z
H.inv <- 1
if (ncol(Z.est) > 1){
H.inv <- varimax(coef(mod.me, type = "matrix")$Z)$rotmat}

# rotate factor loadings
Z.rot <- Z.est %*% H.inv
#extract the rotated loadings and reshape into a matrix
loadings <- t(Z.est) %>% data.frame()
colnames(loadings) <- unique(dat$country)
loadings$x <- c("X1","X2", "X3", "X4")
# add zone to the loadings
loadings <- loadings %>% 
  pivot_longer(-c("x"), names_to = "country", values_to = "loading") %>% 
  mutate(zone = ifelse(country %in% 
    c("Italy","United Kingdom","France","Germany","Spain"),
    "Europe",ifelse( country %in% c ("United States","Canada"), "North America",
    ifelse(country %in% c("Japan", "India", "China"), "Asia",
    ifelse(country %in% c("Argentina","Brazil", "Peru"),"South America", "Oceania")))))


# rotate trends and covnert them to a data frame
trends.rot <- solve(H.inv) %*% mod.me$states
trends.rot <- t(trends.rot)
trends.rot <- data.frame(trends.rot)

# assign column names
colnames(trends.rot) <- c("x1","x2","x3","x4")

# add year and reshape
trends.rot$Year <- seq(from = 1910, to = 2018)
trends.rot <- trends.rot %>% 
  pivot_longer(cols = -c("Year"),
              names_to = "state", values_to = "val") %>% 
  group_by(state) %>% dplyr::arrange(Year, .by_group = TRUE)

# obtain estimated states
s <- tsSmooth(mod.me)
colnames(s) <- c("state", "Year", "val","se")
s$statesRot <- trends.rot$states 
s$Year <- rep(seq(from = 1910, to =2018),4)

```

```{r states, fig.cap = "Smoothed states", code_folding = TRUE}
p1 <- trends.rot %>% ggplot() +
  geom_line(aes(x = Year, y = val, group = state, color = state), size = 1.3, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()+  
  ggtitle("Smoothed states") +
  facet_wrap(~ state ,ncol = 2)


p2 <- s %>% ggplot() +
  geom_line(aes(x = Year, y = val, group = state, color = state), size = 1.3, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()+  
  ggtitle("Rotated states") +
  facet_wrap(~ state,ncol = 2)

p1/p2
```

```{r factors,fig.cap = "Loading factors", code_folding=TRUE}
# Loadings by country
m <- loadings %>% 
  arrange(desc(loading)) %>% 
  ggplot() +
  geom_bar(stat = "identity", aes(x = country, y = loading, 
                                  fill = country, color = country)) +
  scale_color_manual(values = mycolor) +
  scale_fill_manual(values = mycolor) +
  theme_minimal()+  theme(legend.position = "none", panel.background = element_blank(),
                          axis.text.x = element_text(size = 7, angle = 90))+
  ylab("loadings")+
  ggtitle("Factor loadings by country") + 
  xlab("")+
  facet_wrap(~x,ncol = 2)

ggplotly(m)
```

Figure \@ref(fig:states) shows the smoothed states before and after the varimax rotation
and  Figure \@ref(fig:factors) the rotated factors loadings of the different series.
By looking at the loadings in Figure \@ref(fig:factors) 
we immediately see that X1 represent the state  more relevant to all the countries.
All the loadings associated to it are positive and it describes the common upward
trend observed in the exploratory phase. State X3 is characterized by slow negative trend 
followed by a slow increasing ramp at the end at the period. The largest loadings, 
in absolute value associated to it are observed for Italy, Japan, France, china and
India. To conclude state X3 and X4 describe two opposite behaviors. It is interesting
to note that for states X3 and X2 the loadings associated to Italy, India, 
Germany and France follow the exact opposite behavior of those associated to Japan
and China.


## Forecasts
Having inspected the common trends it is time to check if the model is truly able
to produce accurate forecasts for the series considered.
We consider the one step ahead predictions and the associated prediction errors.

```{r, echo = FALSE}
y.hat <- predict(mod.me, type = "ytt1")$pred
y.hat <- y.hat %>% 
  # add columns
  mutate(country = rep(unique(dat$country),each = nrow(y.hat)/15),
  Year = rep(seq(from = 1910, to = 2018), 15))
```


```{r predict, echo = TRUE, fig.caption = "One step ahead predictions", code_folding = TRUE}
p1 <- y.hat %>% ggplot() +
  geom_point(aes(x = Year, y = y, group = country),color = "pink", alpha = 0.7) +
  geom_line(aes(x = Year, y = estimate, group = country), 
            size = 1.2, color = "royalblue", alpha = 0.7) +
  facet_wrap(~country, ncol = 3) +
  ggtitle("One step ahead predictions") +
  theme_minimal()

ggplotly(p1)
```


From plot \@ref(fig:predict) it appears that forecasts produced by the model are 
pretty accurate. Few pronounced errors are observed for the series of Italy and France 
in 1921, 1925 and 1945. In order to quantify the accuracy and compare numerically
the predictive power of the model for the different countries we
compute the following error metrics:

* MSE: $mean((y - \hat{y})^{2})$
* MAE: $mean(|y - \hat{y}|)$
* MAPE:$mean \left( \left| \frac{100( y - \hat{y})}{y} \right|   \right)$

```{r, echo = TRUE}
# define the metrics
mse <- function(x){
   m <- mean((x$y - x$estimate)^2)
   return(m)
}

mae <- function(x){
   m <- mean(abs(x$y - x$estimate))
   return(m)
}

mape <- function(x){
   m <- mean(abs(((100*(x$y - x$estimate))/x$y)))
   return(m)
}

# estimate the metrics for the different series
metrics <- y.hat %>% group_by(country) %>% 
  nest() %>% mutate(mse = lapply(data,mse),
                    mae = lapply(data,mae),
                    mape = lapply(data,mape)) %>% 
unnest() %>% 
  dplyr::select(country,mse,mae,mape) %>% 
  arrange(desc(mse), desc(mae), desc(mape)) %>% 
  unique()

metrics %>% formattable(list(`mae` = color_bar("#FA614B"),
                             `mse` = color_bar("#FA614B"),
                             `mape` = color_bar("#FA614B")))
```
There is a general agreement in the metrics for Italy, France and India that appear
as the countries for which the model produces the worst predictions.
MAjor disagreements are observed for China, Spain and Germany. While according to 
mse and mae the performances in predicting the CO2 emissions of these countries
are not that bad, the same is not observed in the mape. Overall by taking into 
account that no covariates are included in the model and that the states are
modeled as random walk the one step ahead predictions appear good.
However, it is important to note that in order to produce accurate forecasts on 
longer horizons model the states as autoregressive process or including covariates
should be seriously considered.

### Diagnostics
We conclude the analysis with the inspection of the residuals plot in order to 
check if the underlying model assumptions are violated.

```{r, echo = FALSE}
# extract model residuals
mod.res = autoplot(mod.me, plot.type = "model.resids")$data %>% 
  mutate(country = rep(unique(dat$country),each = nrow(y.hat)/15),
         Year = rep(seq(from = 1910, to = 2018), 15))
# extract state residuals
state.res = autoplot(mod.me, plot.type = "state.resids")$data %>% 
  mutate(Year = rep(seq(from = 1911, to =2018),4))
```

```{r fittedres, echo = TRUE, fig.cap = "Fitted residuals", code_folding = TRUE}
# plot model residuals
p1 <- mod.res %>% ggplot() +
  geom_point(aes(x = Year, y = .resids, group = country), color = "royalblue", alpha = 0.4) +
  geom_ribbon(aes(x = Year, ymin = - 1.96*.sigma, 
                  ymax = +1.96*.sigma), alpha = 0.4) +
  facet_wrap(~country, ncol = 3) +
  theme_minimal() +
  ylab("residuals") + ggtitle("Fitted residuals")

ggplotly(p1)
```




```{r stateres, fig.cap = "State residuals", code_foldin = TRUE}
p1 <- state.res %>% ggplot() +
  geom_point(aes(x = Year, y = .resids, group = .rownames), color = "royalblue", alpha = 0.4) +
  geom_ribbon(aes(x = Year, ymin = - 1.96*.sigma, 
                  ymax = +1.96*.sigma), alpha = 0.3) +
  facet_wrap(~.rownames, ncol = 2) +
  theme_minimal() +
  ylab("residuals") + ggtitle("State residuals")

ggplotly(p1)

```

From @\ref(fig:fittedres) we see that the overall fit of the model is not too bad, 
but for some countries such as Italy, France and India the model is not able to 
capture a correlation structure that remains evident in the cyclical pattern of
the residuals. Moreover, some outlying units appear to be present in the data. This is
particularly evident for Italy, France, India and United Kingom.
From figure @\ref(fig:stateres) the state residuals appear well dispersed but
for state X4 there is some autocorrelation evident from the cyclical pattern.


```{r, echo = FALSE}
p1 <- mod.res %>% ggplot() +
  geom_qq(aes(sample = .resids, group = country), color = "royalblue", alpha = 0.4) +
  stat_qq_line(aes(sample = .resids, group = country), color = "red", alpha = 0.4) +
  facet_wrap(~country, ncol = 3) +
  theme_minimal() + 
  ggtitle("State residuals")

ggplotly(p1)
```


```{r, echo = FALSE}
p1 <- state.res %>% ggplot() +
  geom_qq(aes(sample = .resids, group = .rownames), color = "royalblue", alpha = 0.4) +
  stat_qq_line(aes(sample = .resids, group = .rownames), color = "red", alpha = 0.4) +
  facet_wrap(~.rownames, ncol = 2) +
  theme_minimal() + 
  ggtitle("State residuals")

ggplotly(p1)
```
Overall the normality assumption appears satisfied for the fitted residuals, while
in the state residuals a slight deviation is observed in state X3 and X4.


## Conclusions
In this post we have investigated the application of dynamic factor model
to study the CO2 emissions of 15 countries. Despite the model performed well
when the forecast horizon is equal to one year, the inclusion of covariates or 
the use of autoregressive process to describe the evolution of the states could
improve the model performances. This is also confirmed from the
analysis of the residuals that both at observation and state level showed a cyclical
pattern. Moreover the inclusion of dummy variable to account for structural breaks
should be considered in order to improve the validity of the underlying model.

