---
title: "Discovering hidden customer's spending dynamics with Latent Markov Models"
description: |
  In this post we will use Latent Markov Models to model the hidden 
  spending dynamics of a group of customers of a retail company.
  To fit the models we will use the R package LMest.

categories:
  - Econometrics
  - Latent Markov Models
  - Marketing
author: 
  - name: Alessandro Ghiretti
    url: {}
date: 08-01-2021
output: 
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
bibliography: references.bib
nocite: | 
  @WOODSIDE1996189, @bartolucci2012latent, @bartolucci2010overview, @LMest, @bookleeflang2017advanced
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(janitor)
library(kableExtra)
library(LMest)
library(rmarkdown)
library(patchwork)
```



# Introduction
Market or customer segmentation refers to the division of customers or potential customers of a given market into homogeneous groups. Customer segmentation can be performed in different ways, for example a very basic method is to divide customers into groups on the basis of they spending, LTV or number of visits to a store. Where in the first two cases the aim is to identify groups of customers with different profitability and target them with different campaigns, discounts or promotions.

In practice other variables than the number of visits or the spending are taken into account, for example socio-economic variables are often included  as it is well known that factors such as age, lifestyle and status have a great impact on the customer behavior. Once customers are aggregated into different groups the interest might be in monitoring the evolution of these groups over time. For example having collected a sample of customers and having divided them into four groups according to their spending the company would like to know:

* What is the probability that a customer belongs to the a given group?
* What is the probability that an high spender becomes a low spender?
* What is the effect of age on these probabilities?

Latent Markov Models allow to identify hidden states in the data that describe customer behavior. Moreover Latent Markov models provide insights about the dynamics of customers between these hidden states. In the following we will review the theory behind Latent Markov models subsequently apply them to model the customers spending dynamics of a Retailer.

## Latent Markov Models



## Modeling customer spending dynamics via Markov Models
In this section we apply LMM to model the spending dynamics of a sample customers of a Retailer.
We adopt Latent Markov Models for two reasons. The first is that despite common reasoning might be use
to segment customers in different groups depending on their spending we let the data speak and reveal latent groups
that we ignore. The second is that we are aware of the fact that some covariates that might be useful in modeling the
differences in the spending of the customers are not included in the data and therefore we adopt a latent markov model approach to account for unobserved heterogeneity in the data.


### Data description
The dataset that we consider is The Complete Journey dataset freely available from Dunn Humby. According to the description of the data provider the dataset contains household level transactions over two years from a group of 2,500 households who are
frequent shoppers at a retailer. It contains all of each household’s purchases, not just those from a limited number of categories. For certain households, demographic information as well as direct marketing contact
history are included.


### Data analysis
```{r}
setwd("C:\\Users\\USR02193\\OneDrive - Chiesi Farmaceutici S.p.A\\Documents\\Marketing\\Markov")
demo <-  read.csv("hh_demographic.csv")  
transactions <- read.csv("transaction_data.csv") 
```

We start by importing the data into R into two data.frames. demo which contains the demographic information of the customers and, transactions which contains the transactions performed by every customer in the last two years. We use janitor to clean and standardize the name of the columns.

```{r echo = TRUE, eval =TRUE}
# clean column names
colnames(transactions) <- make_clean_names(colnames(transactions))
colnames(demo) <- make_clean_names(colnames(demo))
```

Since we are dealing with temporal data the most natural thing is to retain the information provided by time and apply a longitudinal Latent Markov Model. In order to do so we divide the data into quarters each one composed of 120 days for a total of four quarters per year. 

```{r echo = TRUE, eval =TRUE}
# build the variable indicating the quarter
quarter <- rep(seq(1:8), each = 120)
day <- seq(1:length(quarter))
time_quarter <- data.frame(quarter,day)

# assign the transactions to the corresponding quarter by joining on the day
adjust_transactions <- transactions %>% left_join(time_quarter, by = c("day" = "day"))
```

Subsequently, for each customer we obtain the total spending per quarter which will represent the response variable $Y$ for the LMM.

```{r echo = TRUE, eval =TRUE}
customer_quarter <- adjust_transactions %>% 
  # join with demographics data
  left_join(demo, by = c("household_key" = "household_key")) %>% 
  # group by customer Id
  group_by(household_key, quarter) %>% 
  # obtain total spend per customer
  summarise(tot_spend = sum(sales_value)) %>% 
  ungroup()
```
Finally we bind the obtained quarterly spending with the demographic info and prepare the data for the model.
Before fitting the data we rename the variables, indicating with Ytot_spending the response and with X1variable_name the covariates to be included in the model. We convert all the explanatory variables into factors and remove the customers with missing values or for which the spending in some quarters was not observed. 

```{r, echo = TRUE, eval =TRUE}
# bind demographic info and remove customers with missing values
data_fit <- customer_quarter %>% 
  # join tables
  left_join(demo, by = c("household_key")) %>% 
  # remove households with missing values
  drop_na()

# rename variables for model fitting
c_name <- colnames(data_fit)
nc <- ncol(data_fit)
c_name[3] <- paste("Y",sep ="",colnames(data_fit)[3])
c_name[4:nc] <- paste(paste("X",seq(1:(nc-3)),sep=""),colnames(data_fit)[4:nc],sep = "")
colnames(data_fit) <- c_name

# convert explanatory variables into factors
response <-  colnames(data_fit)[4:nc]
data_fit[response] <- lapply(data_fit[response], factor)

# extract customers which have at least one expense in each quarter
idx = data_fit %>% group_by(household_key) %>% 
  # obtain number of quarters per household
  summarise(n = n()) %>% ungroup() %>%  
  # filter out households
  filter(n==6) %>% 
  # select  hoousehold key and transform it to a vector
  dplyr::select(household_key) %>% as.vector()

data_fit <- data_fit %>% filter(household_key %in% idx$household_key) %>% data.frame()
```

After the cleaning the data comprises 801 customers each of them purcahsed at least one product in eevery quarter.


## Fitting LMM model.

In order to fit the LMM model we will use the LMest ([@LMest]) package, developed by F. Bartolucci, F. Pennoni and .




In order to fit a model with the LMest package we first define a model formula, which
specifies the response variable and the set of covariates affecting the latent states
or the observed variable (see [@LMest] for a full description of the functions of the package).

```{r eval = TRUE, results = "hide"}
#  define formula for model fitting
fmLatent = lmestFormula(data = data_fit, response = "Y", LatentInitial = "X")

# fit the model
mod <- lmestCont(responsesFormula = fmLatent$responsesFormula,
               latentFormula  = fmLatent$latentFormula,
             index = c("household_key","quarter"),
             data = data_fit, 
             paramLatent = "multilogit", k = 1:4)


```

Plot \@ref(fig:mod-sel) shows the  BIC and AIC path for the estimated models assuming different numbers of hidden states. According to it the number of states that best describe the spending dynamics of the customers is 4

```{r mod-sel, fig.cap = 'BIC and AIC paths'}
plot(mod, what = 'modSel')
```


We start by inspecting the prior probabilities, which indicates the probability of a customer to belong to a given state, and the corresponding means.

```{r}
rbind(mod$Mu)
```
The four latent states correspond to four different customer segments.
Low spenders, middle-low spenders, middle-high spenders and high spenders.
By looking at the path diagram it appears that it is particularly difficult
that a customer switches between the di
```{r, figures-side, fig.show = 'hold'}
 plot(mod1, what ='transitions', main = "M2 Averaged transition probabilities")
```
The plots represent the Markov diagram and the associated transition probabilities. As model M2 allows for time varying probabilities, the diagram shows the averaged transition probabilities over the 6 quarters. We will focus on the interpretation of the plot describing the M2 dynamics. From the diagram it is evident that high spender customers have the highest probability of remaining in their actual state over the next 4 months. This is a really positive news for the company as it might represent the fact that the company is retaining is high segment properly. The same also holds for middle-high spenders that have an estimated probability of 70% of remaining in the state. However, in this case there in an average estimated probability of 14% that a middle-spender customer will become a middle-low spender over the next quarter. Moreover, the estimated probability of a low spender to increase his spending and become a middle spender is 17% which is really positive since once a customer becomes a middle-low spender the probability of subsequently to state 3 and increase further his spending is equal to 20%.

The dynamics of the customers between the different ghroups can be investingated by looking ad the marginal distributions over time.

```{r prob-dist, fig.cap = "Marginal distributions"}
 g  = plot(mod1, what ='marginal')
```
Fig \@ref(fig:prob-dist) shows the distribution of the states over time. The high 
spenders segment shows the highest stability over the 6 quarters.
The low spenders segment show a steady decline between the first and the second quarter with
an increase in the number of subjects belonging to the middle groups.
there is an increase in the number of customers belonging to the middle spenders groups with a 
slight decline between the fifth and sixth quarter.

### LMM with covariates
We conclude the analysis discussing the panel Latent Markov model fitted including the
explanatory variables. 

